A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-08-15 paper/comments/comments.tex
included some sample drawings and mentioned parameter settings in the text

\tcomment{l.158: How do you choose $\rho$?}

\tcomment{l.165: You should discuss the choice of the parameters $\delta_{\min}$ and $\delta_{\max} $.}

>>>>

\rcomment{l.158: How do you choose $\rho$?}

\response{Added the following comment: For our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$ (where $w$ and $h$ are the width and the height of the initial drawing, respectively), $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=50$.\footnote{In order to further decrease the time needed for an iteration, we only computed a candidate position for 10 of the 50 rays.} }

\rcomment{l.165: You should discuss the choice of the parameters $\delta_{\min}$ and $\delta_{\max} $.}

\response{See above.}

### 2018-08-15 paper/paper.tex
included some sample drawings and mentioned parameter settings in the text

@@

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works.
%In Section~\ref{sec:preliminaries}, we introduce preliminary notions and the notation used throughout the paper.
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. In the appendix, we provide experimental results on grid restricted drawings, on more test sets and on the graphs from the Graph Drawing Competition in 2017.

@@

% ==================================================================
\myparagraph{Grid drawings}
% ==================================================================
%
Our algorithm, as it has been described so far, does not necessarily produce grid drawings, i.e., drawings in which the vertices are at integer coordinates. However, it can be easily adjusted to produce such drawings. More precisely, if we round the candidate positions computed at each iteration of our algorithm to their closest grid points and use these grid points as candidates for the next position of the vertex to be moved, then the obtained drawing will be grid (assuming, of course, that the starting drawing is grid). One can even bound the size of the grid, by rejecting candidate grid positions outside the bounds. In Appendix~\ref{app:grid}, we report experimental results on this variant.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is available on request from the authors.} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; in Appendix~\ref{app:north}, we also report on the AT\&T graphs.

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}

@@

Regarding the number of crossings, we observe that the restricted variant of our algorithm and the force-directed algorithm of Huang et al.\ yield drawings with comparable number of crossings, which at the same time is significantly smaller than the corresponding number of crossings produced by the two other algorithms of our experiment; refer to Fig.~\ref{fig:cr-res-3}.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings, in which the crossing angles, are large are easy to read and understand. What we observed is that the drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other (nearly at the same spot). Hence, the angular resolution and the aspect ratio were very often poor. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.

We conclude by noting that our motivation to work with this problem was our participation to GD2017 contest, where we performed miserably using a force-directed algorithm; for details see Appendix~\ref{app:contest}. As our evaluation shows, the performance of such algorithms is good, only when several aesthetic criteria are taken into account; our new approach is definitely more promising than our previous (as also evidenced by our experimental evaluation). The framework that we developed seems to be quite adaptable to optimize or to take into account also other desired aesthetic properties of a drawing (even though our initial plan was to optimize only the crossing resolution).

@@

\newpage
\appendix

@@

>>>>

%conf for conference version, arxiv for arxiv version
\newcommand{\ver}{arxiv}
\newcommand{\arxapp}[2]{\ifthenelse{\equal{\ver}{conf}}{#2}{#1}}

@@

\medskip\noindent{\it Structure of the paper:}
The remainder of this paper is structured as follows. Section~\ref{sec:relatedwork} overviews related works.
%In Section~\ref{sec:preliminaries}, we introduce preliminary notions and the notation used throughout the paper.
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones of Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} and Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. \arxapp{In the appendix}{In~\cite{arxivVersion}}, we provide experimental results on grid restricted drawings, on more test sets and on the graphs from the Graph Drawing Competition in 2017.

@@

% ==================================================================
\myparagraph{Grid drawings}
% ==================================================================
%
Our algorithm, as it has been described so far, does not necessarily produce grid drawings, i.e., drawings in which the vertices are at integer coordinates. However, it can be easily adjusted to produce such drawings. More precisely, if we round the candidate positions computed at each iteration of our algorithm to their closest grid points and use these grid points as candidates for the next position of the vertex to be moved, then the obtained drawing will be grid (assuming, of course, that the starting drawing is grid). One can even bound the size of the grid, by rejecting candidate grid positions outside the bounds. \arxapp{In Appendix~\ref{app:grid}}{In~\cite{arxivVersion}}, we report experimental results on this variant.

@@

In this section, we present the results of our experimental evaluation. For comparison purposes, apart from our algorithm, we also implemented the force-directed algorithms of Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13}. The implementations\footnote{Our implementation is available on request from the authors.} were in Java using yFiles~\cite{DBLP:books/sp/04/WieseE004}. The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM.
%
As a test set for our experiment, we used the non-planar Rome graphs~\cite{DBLP:reference/crc/BattistaD13}, which form a collection of around 8.100 benchmark graphs; \arxapp{in Appendix~\ref{app:north}}{in~\cite{arxivVersion}}, we also report on the AT\&T graphs.

The experiment was performed as follows. Initially, each Rome graph was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with
\begin{inparaenum}[(i)]
\item our algorithm,
\item our algorithm restricted~not to violate the aspect ratio of the initial layout, and the force-directed algorithms
\item by Argyriou et al.\ and
\item by Huang et al.
\end{inparaenum}
For our algorithm, we chose $\delta_{max}=\frac{1}{2}\max\{w,h\}$ (where $w$ and $h$ are the width and the height of the initial drawing, respectively), $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=50$.\footnote{In order to further decrease the time needed for an iteration, we only computed a candidate position for 10 of the 50 rays.}

@@

We observe that the restricted variant of our algorithm and the force-directed algorithm of Huang et al.\ yield drawings with comparable number of crossings, which at the same time is significantly smaller than the corresponding number of crossings produced by the two other algorithms; refer to Fig.~\ref{fig:cr-res-3}.

@@

% ==================================================================
\myparagraph{Discussion}
%
% ==================================================================
%
%In this paper, we introduced a new heuristic aiming to produce drawings of high crossing resolution, for which we also presented variants that take into account other common aesthetic criteria in Graph Drawing. Our experimental evaluation indicates that the new heuristic is competitive to state of the art force-directed algorithms, even when restricted to a given maximum aspect ratio. As a future direction, we plan to evaluate the performance of variants of our heuristic that compromise between even more aesthetic criteria.
%
While working on this project, we made some useful observations and  obtained some interesting insights. In particular, there is a recent hypothesis (also supported by experiments) that drawings, in which the crossing angles, are large are easy to read and understand. We observed that drawings that are optimized only in terms of the crossing angles might be arbitrarily bad and may have several undesired properties. In particular, in these drawings it was very common to have adjacent edges to run almost in parallel and vertices to be very close to each other. Hence, angular resolution and aspect ratio were often poor. The additional restrictions that we imposed regarding the angular resolution and the aspect ratio helped significantly improving the readability of the drawings, without loosing too much of their quality in terms of the crossing resolution.

We conclude by noting that our motivation to work with this problem was our participation to GD2017 contest, where we performed miserably using a force-directed algorithm; for details \arxapp{see Appendix~\ref{app:contest}}{see~\cite{arxivVersion}}. As our evaluation shows, the performance of such algorithms is good, only when several aesthetic criteria are taken into account; our new approach is definitely more promising than our previous as evidenced by our experiments. The framework that we developed seems to be quite adaptable to optimize or to take into account also other desired aesthetic properties of a drawing.

@@

\arxapp{\newpage
\appendix

@@

\newpage

\section{Sample Drawings}
\label{app:samples}

\begin{figure}[htbp]
\centering
%\subfloat[\label{fig:graph5-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph5-yfiles}
%}

\subfloat[\label{fig:graph5-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-angular}
}
\hfill
\subfloat[\label{fig:graph5-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-ourBest-total}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph5-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph5-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-yfiles}
}
\hfill
\subfloat[\label{fig:graph5-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph5-huang-total}
}
\caption{Drawings of graph 5 from last years graph drawing contest.  
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~Our results, 
(d)--(f)~Results computed with Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, 
(g)--(i)~Results computed with Huang et al~\cite{DBLP:journals/vlc/HuangEHL13}.}
\label{fig:graph5}
\end{figure}

\begin{figure}[htbp]
\centering
%\subfloat[\label{fig:graph9-yfiles}{Organic Layout}]{
%\centering
%\includegraphics[width=0.3\textwidth]{graph9-yfiles}
%}

\subfloat[\label{fig:graph9-ourBest-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-ourBest-crossing}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-ourBest-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-crossing}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-angular}
}
\hfill
\subfloat[\label{fig:graph9-argyriou-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-argyriou-total}
}
\hfill
\subfloat[\label{fig:graph9-huang-cr}{Crossing Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-ar}{Angular Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-yfiles}
}
\hfill
\subfloat[\label{fig:graph9-huang-tr}{Total Resolution}]{
\centering
\includegraphics[width=0.3\textwidth]{graph9-huang-total}
}
\caption{Drawings of graph 9 from last years graph drawing contest. 
%(a)~Organic Layout by \emph{yFiles}, 
(a)--(c)~Our results, 
(d)--(f)~Results computed with Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13}, 
(g)--(i)~Results computed with Huang et al~\cite{DBLP:journals/vlc/HuangEHL13}.}
\label{fig:graph9}
\end{figure}}{}
